---
title: "Thanks for Reviewing Bug Reports"
author: "Tomas Kalibera, Luke Tierney"
date: 2019-12-16
categories: ["External Contributions"]
tags: ["debugging", "bugs", "bugzilla"]

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Just a quick update and note of thanks to all who have responded to
our [blog post](https://blog.r-project.org/2019/10/09/r-can-use-your-help-reviewing-bug-reports/index.html)
on 2019-10-09 and helped with reviewing and resolving bug reports.
Thanks to your help the pace at which bugs have been resolved has
increased nicely since the date of that post:

```{r, echo = FALSE, message = FALSE}
library(tidyr)
library(dplyr)

load("2019-12-16-reviewing-bugs-thanks.Rdata")
dd <- select(d, id, created, closed, status)
dd <- filter(dd, status != "CLOSED" | ! is.na(closed))
ddl <- gather(dd, event, time, created, closed)
ddl <- arrange(ddl, time)

ddl$nclosed = cumsum(ddl$event == "closed")
plot(nclosed ~ time, data = filter(ddl, time >= "2019-01-01"), type = "l",
     ylab = "Cumulative Number of Bugs Closed", xlab = "Date")
abline(v = as.POSIXct("2019-10-09"), lty = 2)
axis(3, at = as.POSIXct("2019-10-09"), labels = "2019-10-09")
```

Our special thanks to Michael Chirico, Henrik Bengtsson, Suharto Anggono,
Hugo Gabriel Eyherabide and Elin Waring, who have contributed most in this
period.  There is still a a lot of work to do, so please keep your
contributions coming!







