---
title: "New Tricks for the R Sweave Drivers"
author: "Vincent Goulet and Kurt Hornik"
date: "2026-09-18"
categories: ["R Core"]
tags: ["Sweave", "RweaveLatex", "Rtangle"]

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Introduction

The concept of literate programming was pioneered by Donald Knuth in
1984 to improve the documentation of computer programs, as this famous
quote explains:

> I believe that the time is ripe for significantly better
> documentation of programs, and that we can best achieve this by
> considering programs to be *works of literature*. Hence, my
> title: "Literate Programming."

The aim of literate programming is to create programs that are
suitable literature for *human beings* rather than a set of
instructions to a computer. A program becomes a web of two languages:
a document formatting language, and a programming language. In Knuth's
original presentation, these were respectively TeX and Pascal. A
program, or a set of programs, then process this web of code. Knuth
developed the [WEB](https://en.wikipedia.org/wiki/WEB) system and
coined the terms *weaving* for the process of creating the
documentation, and *tangling* for the process of producing the
executable program.

Friedrich Leisch brought literate programming to the R ecosystem with
the [Sweave](https://en.wikipedia.org/wiki/Sweave) framework. Based on
the system [noweb](https://en.wikipedia.org/wiki/Noweb), Sweave
allowed to combine text in LaTeX and R code for literate statistical
practice and reporting. This sparkled the use of *vignettes* for
improved documentation within R and contributed packages, and it
certainly contributed to the rise of reproducible research practices
in statistical computing. Sweave was also of course the inspiration
for [knitr](https://en.wikipedia.org/wiki/Knitr), by far the most
widely used literate programming system in the R community nowadays,
if only as the basis of the document format R Markdown.

Although knitr provides more bells and whistles, Sweave remains
simpler and, perhaps most importantly for the forthcoming discussion,
it continues to give equal treatment to the lesser known of the two
literate programming procedures: tangling.

Sweave was designed from the outset in a modular way to allow
different drivers for the weaving and tangling procedures. The
standard drivers are `RweaveLatex` to transform `.Rnw` files with
LaTeX documentation chunks and R code into `.tex` files, and `Rtangle`
to extract R code from a `.Rnw` file into `.R` scripts. A number of
packages propose additional drivers for Sweave, for example
[ascii](https://cran.r-project.org/package=ascii),
[highlight](https://cran.r-project.org/package=highlight) or
[R2HTML](https://cran.r-project.org/package=R2HTML), just to name a
few; see also the [*Reproducible
Research*](https://CRAN.R-project.org/view=ReproducibleResearch) CRAN
Task View.

The feature set of Sweave has remained remarkably stable since its
inception. This blog introduces new tricks for the standard drivers
that we added in version 4.6.0 of R. Many of these first came to life
in the now deprecated package
[RweaveExtra](https://cran.r-project.org/package=RweaveExtra).
	
# Motivation

The standard Sweave drivers of package utils used to enforce some
conventions that, if sensible in a statistical analysis context, could
prove limiting in more general uses of literate programming. We will
discuss the following conventions in the sequel:

1. The values of chunk options have to be given explicitly, with no
   possibility to reuse values computed in earlier code chunks.
2. Expressions in code chunks are always parsed on weaving, whether
   the option `eval` is `TRUE` or `FALSE`.
3. Code chunks with `eval=FALSE` are automatically and irrevocably
   commented out on tangling.
4. Tangling allows to control whether evaluated code chunks are
   extracted to a single file or to separate files via the option
   `split`, but not whether or not a given code chunk is extracted in
   the first place (the option `drop.evalFALSE=TRUE` only omits
   unevaluated chunks).
5. Code chunks collected in the same file on tangling end up always
   separated by two blank lines, and these are also present at the end
   of the file.
6. The names of tangled files always end with a `.R` extension
   (omitting the other supported default of `.S` that is probably not
   much used nowadays).

The first item above limits the amount of logic one can build into a
literate programming document. The combination of the second and third
items makes it impossible to include in a code chunk invalid R code
that should appear uncommented in the tangled script. The fourth and
fifth items illustrate that authors have little control over what is
extracted on tangling, and in what format. Finally, the sixth item
becomes a limitation once the parsing requirement of item 2 is lifted
and code chunks may contain code in languages other than R entirely.

# Objects as chunk options

In Sweave, code chunks start with `<<`*options*`>>=` at the beginning
of a line, where the optional *options* have the form `key`=`value`
and are separated by commas. All options must take a value that used
to be either logical, numerical, or character.

The `RweaveLatex` driver now allows the value of logical and numerical
chunk options (only) to be the name of an object defined in earlier,
evaluated code chunks. This is useful to pass computed values to
options. The example below uses this feature to evaluate a chunk only
if a package is available, and to create a plot with computed
dimensions (this example and all others in the sequel are stripped of
all non-essential text content).

```
<<>>=
hasfoo <- requireNamespace("foo", quietly = TRUE)
ht <- 5
wd <- (1 + sqrt(5))/2 * ht
@

<<eval=hasfoo>>=
foo::foo(42)
@

<<echo=FALSE, fig=TRUE, width=wd, height=ht>>=
data(airquality, package="datasets")
library("graphics")
boxplot(Ozone ~ Month, data = airquality)
@
```

This type of option evaluation allows Sweave package vignettes to
conveniently deal with the situation that suggested dependencies (that
is: mentioned in the `Suggests` field of the `DESCRIPTION` file) are
used conditionally when checking.

Why limit the feature to logical and numerical options? All the values
of options first reach the Sweave driver as character strings. The
driver then tries to coerce them to the expected type. When that
fails, it is fairly safe for logical and numerical options to let the
driver check if an object of the correct mode exists in the
environment, and use its value when that is the case. Trying to do the
same for character options would prove impossible, or at least very
risky, since the driver could not tell apart intended values of
options from object names. Given that the vast majority of the options
of the `RweaveLatex` driver are either logical or numerical, we did
not deem necessary to devise a scheme to extend the feature to
character options.

# Unparsable code in code chunks

Automatic parsing of code chunks is a nifty Sweave feature: it ensures
that all code in a document is at least syntactically valid. However,
sometimes unparsable code is a feature, not a bug. For example, an
instructor may wish to include erroneous R code in a chunk for
didactic purposes, yet distribute the tangled code uncommented such
that learners may nevertheless evaluate it.

To cope with such situations, the `RweaveLatex` driver has gained
options to completely ignore code chunks on weaving. Chunks using
`ignore.on.weave=TRUE` are not parsed on weaving, yet remain written
out verbatim on tangling. A shorter alias `weave` for
`!ignore.on.weave` is also supported. An option `ignore` sets at once
`ignore.on.weave` and the equivalent option for the `Rtangle` driver
(see below).

The following example illustrates the use of the new option to
maintain a document that evaluates without error on weaving, yet
produces an uncommented script on tangling.

```
<<>>=
## Some R comparison operators.
486 < 521                  # smaller than
486 >= 521                 # greater or equal
486 != 521                 # different from
@
<<ignore.on.weave=TRUE>>=
## Watch out! Equality is not tested with '='.
5 = 2                      # syntax error
@
<<>>=
## The equality comparison operator is '=='.
486 == 486                 # equality
@
```

We will cover another interesting use of this option in the following
section.

# Finer control of the tangling process

New options for the `Rtangle` driver address the last four motivations
mentioned earlier, thereby giving users finer control on the tangling
process.

First, `Rtangle` also gains an option `ignore.on.tangle` to completely
omit a code chunk on tangling, irrespective of the value of option
`eval`. The driver also supports the shorter alias `tangle` for
`!ignore.on.tangle`, and the option `ignore` to set both
`ignore.on.weave` and `ignore.on.tangle` at once. The possibility to
skip a chunk on tangling is convenient to avoid cluttering the tangled
script with unimportant code or, when using `split=TRUE`, the file
system with unneeded scripts.

Second, the new option `chunk.sep` allows to specify the separator
between code chunks when they are collected in a single file. The
historic default of two blank lines is maintained. One may also use
`chunk.sep=FALSE` to completely omit the separator, either throughout
a script when used as an argument to the `Rtangle` call, or between
two chunks when used as a chunk option. It is also worth noting that
the chunk separator is now added *before* all chunks, except the first
one. Therefore, there is no longer a chunk separator at the end of
tangled scripts.

Finally, `Rtangle` now provides an option `extension` to specify the
extension, without the leading dot, for the file name of a tangled
code chunk when splitting is selected. If used as `extension=TRUE`,
the default extension (usually `R`) is used. If the option is `FALSE`,
no extension is added to the file name. The option `extension` is
specially useful with `ignore.on.weave=TRUE` of `RweaveLatex` to
include code or text that the engine would not be able to parse on
weaving, yet that needs to be tangled to a file with a specific
extension, such as `.sh` or `.txt`.

Consider a file `example.Rnw` that contains the following code chunks:

```
<<script, echo=TRUE>>=
1:10
@

<<echo=FALSE, ignore.on.tangle=TRUE>>=
x <- 42
@

<<script>>=
gamma(1:5)                 # gamma function
factorial(0:4)             # factorial
@
<<script, ignore.on.weave=TRUE, chunk.sep=FALSE>>=
?gamma                     # all related functions
@

<<hello, extension=sh, ignore.on.weave=TRUE>>=
#!/bin/sh

echo "Hello, World!"

exit 0
@

<<README, extension=FALSE, ignore.on.weave=TRUE>>=
Hello, World!
@
```

We use `ignore.on.weave=TRUE` in the fourth chunk to hide the help
system call from the weaving process. Now, tangling the file with the
following expression will create three files: `script.R`, `hello.sh`,
and `README`.
```{r, echo=TRUE, eval=FALSE}
Stangle("example.Rnw", split = TRUE, prefix = FALSE,
         annotate = FALSE, chunk.sep = "\n")
```

The file `script.R` is shown below. Notice that: the expression `x <-
42` is omitted from the script; the help system call is present and
uncommented; the two main blocks of code are separated by a single
blank line; what were originally two code chunks became only one
block; there are no blank lines at the end of the file.

```
--- script.R ---
1:10

gamma(1:5)                 # gamma function
factorial(0:4)             # factorial
?gamma                     # all related functions
```

As for `hello.sh` and `README`, shown below, the fact that the files
exist with an extension other than `.R`, and that their contents is
*not* R code, illustrates how one may use the options `extension` and
`ignore.on.weave` to maintain various pieces of code or text inside a
single source file.
```
--- hello.sh ---
#!/bin/sh

echo "Hello, World!"

exit 0
```

```
--- README ---
Hello, World!
```

# Obtaining the name of the processed file

Weaving and tangling are often seen as separate, somewhat orthogonal
procedures. However, one may devise various schemes where both weaving
*and* tangling are needed to create a document. Here are a few examples:

- maintaining documentation and code together even though the code
  is not directly used to create the text;
- code that needs to be included verbatim in a document, but not
  necessarily as or where it appears in the sources;
- pieces of code that rely on other pieces being saved as files.

Vincent (the first author) proposes in a [TUGboat article](https://doi.org/10.47397/tb/47-1/tb145goulet-stangle)
a nice strategy to achieve this: embedding a tangling procedure 
*inside* a weaving procedure by calling `Stangle` in a code chunk. 
Now, this requires the name of the file to process... which is the 
file being processed. To avoid hard coding the file name in the
`Stangle` call, we also introduced in the Sweave framework the 
function `SweaveGetSourceName` that returns the name of the file being
woven by an `Sweave` process launched from the command line. This 
allows to start `Stangle` inside `Sweave` with the a code chunk such as:
```{r, echo=TRUE, eval=FALSE}
<<echo=FALSE, results=hide, ignore.on.tangle=TRUE>>=
FILE <- SweaveGetSourceName()
Stangle(FILE)
@
```

The function works by parsing the output of `commandArgs`, and
supports weaving processes started by means of either `R CMD Sweave`,
`R -e`, or `Rscript -e`.
